---
title: "Business Statistics Using R - Individual Assignment"
author: 'Aastha Bajracharya'
date: "20th August 2026"
output:
  html_document:
    toc: false
  pdf_document:
    toc: false
  word_document:
    toc: false
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,
                       fig.width = 8, fig.height = 4.5)
```

# Introduction

>This analysis investigates factors influencing customer expenditure in an
e-commerce environment. The study applies exploratory data analysis,
descriptive statistics, hypothesis testing, and regression modelling to 
identify significant predictors of spending behavior and customer retention.

# 1. Environment Setup & Data Ingestion

## 1.1 Loading Library

```{r}
library(tidyverse)
library(ggplot2)
library(car)
library(corrplot)
library(gridExtra)
library(janitor)
library(broom)
library(knitr)
```

## 1.2 Loading the Raw Dataset

```{r}
df_raw <- read.csv("ecommerce_customer_behavior_dataset.csv")
```

# 2. Checking Data Quality

## 2.1 Initial Data Quality Check

```{r}
str(df_raw)
summary(df_raw)
colSums(is.na(df_raw))
```

```{r, echo = FALSE}
n_na  <- sum(is.na(df_raw))
n_dup <- sum(duplicated(df_raw))
```

>**Findings:** The data set has `r nrow(df_raw)` rows and `r ncol(df_raw)`
columns, with **`r n_na` missing values** and **`r n_dup` duplicate rows**.

## 2.2 Column Standardization

```{r, echo=TRUE}
df <- df_raw %>%
  clean_names() %>%
  distinct() %>%
  mutate(
    date = as.Date(date),
    gender = as.factor(gender),
    city = as.factor(city),
    product_category = as.factor(product_category),
    payment_method = as.factor(payment_method),
    device_type = as.factor(device_type),
    is_returning_customer = as.factor(is_returning_customer)
  )
```

## 2.3 IQR Method for Outlier Detection 

```{r, echo=FALSE}
numeric_cols <- c("total_amount", "unit_price", "quantity", "discount_amount",
                  "age", "session_duration_minutes")

outlier_summary <- map_dfr(numeric_cols, function(v) {
  x <- df[[v]]
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower <- q1 - 1.5 * iqr
  upper <- q3 + 1.5 * iqr
  n_out <- sum(x < lower | x > upper, na.rm = TRUE)
  tibble(Variable = v, Lower_Bound = round(lower, 2), Upper_Bound = round(upper, 2),
         N_Outliers = n_out, Pct_Outliers = round(100 * n_out / length(x), 1))
})

# Extract total_amount metrics for the inline text below
ta_info <- outlier_summary %>% filter(Variable == "total_amount")
upper_bound <- ta_info$Upper_Bound
n_outliers  <- ta_info$N_Outliers

kable(outlier_summary, caption = "IQR Outlier Check Across All Numeric Variables")
```

>`discount_amount` has the highest outlier rate (17.6%) since most orders
carry no discount at all, making any real discount register as an
"outlier" by the IQR rule. `quantity` has none, its range is naturally
narrow (1-6 units).  Using the standard IQR rule, `total_amount` values above **$`r round(upper_bound, 2)`** are
flagged as potential outliers — `r n_outliers` orders
(`r round(100 * n_outliers / nrow(df), 1)`%) exceed this threshold.

>**Decision:** The outliers are not removed. They could correspond to genuine 
high-value orders rather than data-entry errors, and removing real 
revenue-generating transactions would bias the analysis. Instead, the 
right-skew this causes is addressed with a log-transformation of the target variable.

## 2.4 Visualization of Original Right-Skewed Distribution

```{r, echo =FALSE}
ggplot(df, aes(x = total_amount)) +
  geom_histogram(bins = 30, fill = "#1F78B4", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(total_amount)),
             color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(aes(xintercept = median(total_amount)),
             color = "darkgreen", linetype = "dotdash", linewidth = 1) +
  labs(title = "Distribution of Total Expenditure",
       subtitle = "Red = Mean | Green = Median",
       x = "Total Amount ($)", y = "Count") +
  theme_minimal()
```

>The distribution is strongly right-skewed: the mean sits well above the
median, pulled up by the high-value orders identified above.

## 2.5 Log Transformation

```{r, echo =FALSE}
df <- df %>%
  mutate(log_total_amount = log(total_amount))

p1 <- ggplot(df, aes(x = total_amount)) +
  geom_histogram(bins = 30, fill = "#1F78B4", color = "black", alpha = 0.7) +
  labs(title = "Raw Total Amount", subtitle = "Severe right skew",
       x = "Total Amount ($)", y = "Count") +
  theme_minimal()

p2 <- ggplot(df, aes(x = log_total_amount)) +
  geom_histogram(bins = 30, fill = "#33A02C", color = "black", alpha = 0.7) +
  labs(title = "Log(Total Amount)", subtitle = "Approximately symmetric",
       x = "log(Total Amount)", y = "Count") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

>**Why this step was done:** The raw target variable violates linear
regression's normality/constant-variance assumptions and the log transform is
the standard fix, hence a log-based model is fitted alongside the raw-scale
ones.

# 3. Exploratory Data Analysis (EDA)

## 3.1 Data Types

```{r, echo= FALSE}
glimpse(df)
```

>The dataset mixes numeric variables (`total_amount`, `unit_price`,
`quantity`, `discount_amount`, `age`, `session_duration_minutes`),
categorical/factor variables (`gender`, `city`, `product_category`,
`payment_method`, `device_type`, `is_returning_customer`), and a date
variable (`date`).

## 3.2 Descriptive Statistics

```{r, echo = FALSE}
skewness <- function(x) {
  n <- length(x)
  (sum((x - mean(x))^3) / n) / (sd(x)^3)
}

descriptive_table <- tibble(
  Variable = c("Total Amount", "Unit Price", "Quantity", "Discount Amount",
               "Age", "Session Duration (min)"),
  Mean     = c(mean(df$total_amount), mean(df$unit_price), mean(df$quantity),
               mean(df$discount_amount), mean(df$age),
               mean(df$session_duration_minutes)),
  Median   = c(median(df$total_amount), median(df$unit_price), median(df$quantity),
               median(df$discount_amount), median(df$age),
               median(df$session_duration_minutes)),
  SD       = c(sd(df$total_amount), sd(df$unit_price), sd(df$quantity),
               sd(df$discount_amount), sd(df$age),
               sd(df$session_duration_minutes)),
  IQR      = c(IQR(df$total_amount), IQR(df$unit_price), IQR(df$quantity),
               IQR(df$discount_amount), IQR(df$age),
               IQR(df$session_duration_minutes)),
  Skewness = c(skewness(df$total_amount), skewness(df$unit_price), skewness(df$quantity),
               skewness(df$discount_amount), skewness(df$age),
               skewness(df$session_duration_minutes))
) %>%
  mutate(across(where(is.numeric), ~round(.x, 2)))

kable(descriptive_table,
      caption = "Descriptive Statistics")
```

>`Total_Amount` shows strong positive skew along with
`Age` and `Session_Duration_Minutes` being close to symmetric.

## 3.3 Univariate Analysis

```{r, echo = FALSE}
ggplot(df, aes(x = age)) +
  geom_histogram(bins = 20, fill = "#6A3D9A", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Customer Age", x = "Age", y = "Count") +
  theme_minimal()
```

>Customer age is fairly evenly spread across the adult range, with no strong
skew.

## 3.4 Bivariate Analysis: Spend by Product Category

```{r, echo=FALSE}
ggplot(df, aes(x = product_category, y = total_amount, fill = product_category)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Customer Expenditure across Product Categories",
       x = "Product Category", y = "Total Amount ($)") +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r, echo=FALSE}
df %>%
  group_by(product_category) %>%
  summarise(
    Mean_Spend = mean(total_amount),
    Median_Spend = median(total_amount)
  ) %>%
  arrange(desc(Mean_Spend)) %>%
  kable(
    caption = "Average Customer Expenditure by Product Category"
  )
```
>Electronics records the highest average expenditure, while Books and Beauty generate the lowest average order values. This supports the boxplot findings that spending patterns vary substantially across product categories.

## 3.5 Bivariate Analysis: Unit Price vs Total Amount

```{r, echo=FALSE}
ggplot(df, aes(x = unit_price, y = total_amount)) +
  geom_point(alpha = 0.3, color = "#1F78B4") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Unit Price vs Total Expenditure",
       x = "Unit Price ($)", y = "Total Amount ($)") +
  theme_minimal()
```

>A clear positive linear relationship, though the spread widens at higher
prices, orders with expensive items also vary more in quantity bought.

## 3.6 Bivariate Analysis: Quantity vs Total Amount
```{r, echo=FALSE}
ggplot(df,
       aes(x = quantity,
           y = total_amount)) +
  geom_point(alpha = 0.3,
             color = "#1F78B4") +
  geom_smooth(method = "lm",
              color = "red") +
  labs(
    title = "Quantity Purchased vs Total Expenditure",
    x = "Quantity",
    y = "Total Amount ($)"
  ) +
  theme_minimal()
```

>The steepest, most consistent relationship of the three bivariate plots,
matches quantity's status as the strongest driver in the regression models.

## 3.7 Univariate and Bivariate Analysis: Discount Amount
```{r, echo=FALSE}
pct_zero_discount <- round(100 * mean(df$discount_amount == 0), 1)

ggplot(df, aes(x = discount_amount)) +
  geom_histogram(bins = 30, fill = "#E31A1C", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Discount Amount",
       subtitle = paste0(pct_zero_discount, "% of orders receive no discount at all"),
       x = "Discount Amount ($)", y = "Count") +
  theme_minimal()
```

`r pct_zero_discount`% of orders in this dataset have a discount amount
of exactly $0. Among the remaining orders, the distribution is right-skewed,
in the same direction as total amount and unit price, though far more
zero-inflated than either.

```{r, echo=FALSE}
ggplot(df, aes(x = discount_amount, y = total_amount)) +
  geom_point(alpha = 0.3, color = "#E31A1C") +
  geom_smooth(method = "lm", color = "black") +
  labs(title = "Discount Amount vs Total Expenditure",
       x = "Discount Amount ($)", y = "Total Amount ($)") +
  theme_minimal()
```

```{r, echo=FALSE}
raw_cor_discount <- round(cor(df$discount_amount, df$total_amount), 3)
```
## 3.8 Multivariate Analysis: Correlation Matrix

```{r, echo= FALSE}
numeric_vars <- df %>% select(total_amount, unit_price, quantity, discount_amount, age, session_duration_minutes)
cor_matrix <- cor(numeric_vars)
corrplot(cor_matrix, method = "number", type = "upper", tl.col = "black")
```

>`unit_price` and `quantity` show the strongest correlation with
`total_amount`; `age` and `session_duration_minutes` are essentially
uncorrelated with spend.

# 4. Regression Modelling

## 4.1 Building the Models

>Four models are compared, moving from a simple one-predictor model
to a full model with demographic and category controls, plus a
log-transformed version of the full model. These test hypotheses H1-H4:

- **H1 (H0: β = 0):** Order quantity has a significant relationship with total spend.
- **H2 (H0: β = 0):** Unit price has a significant relationship with total spend.
- **H3 (H0: β = 0):** Customer age has a significant relationship with total spend.
- **H4 (H0: β = 0):** Discount amount has a significant relationship with total spend.
- **H5 (H0: β = 0):** Total customer expenditure varies significantly across product categories.

```{r, echo=TRUE}
# Model 1: Simple Linear Regression (Order Quantity as primary driver)
m1_slr <- lm(total_amount ~ quantity, data = df)

# Model 2: Multiple Linear Regression (Pricing & Promotion focus)
m2_mlr_pricing <- lm(total_amount ~ unit_price + discount_amount, data = df)

# Model 3: Full Multiple Linear Regression (Behavioral & Demographics)
m3_mlr_full <- lm(total_amount ~ unit_price + quantity + discount_amount + age + product_category, data = df)

# Model 4: Full Multiple Linear Regression Model using Log-Transformed Total Amount
m4_mlr_full_log <- lm(log_total_amount ~ unit_price + quantity + discount_amount + age + product_category, data = df)
```

```{r, echo=FALSE}
kable(tidy(m1_slr) %>% select(term, estimate, std.error, statistic, p.value),
      digits = 4, caption = "Model 1: Total Amount ~ Quantity")
kable(tidy(m2_mlr_pricing) %>% select(term, estimate, std.error, statistic, p.value),
      digits = 4, caption = "Model 2: Total Amount ~ Unit Price + Discount Amount")
kable(tidy(m3_mlr_full) %>% select(term, estimate, std.error, statistic, p.value),
      digits = 4, caption = "Model 3: Full model (raw scale)")
kable(tidy(m4_mlr_full_log) %>% select(term, estimate, std.error, statistic, p.value),
      digits = 4, caption = "Model 4: Full model (log-transformed Total Amount)")
```

## 4.2 Overall Effect of Product Category

`product_category` enters Model 4 as 7 separate dummy coefficients, each
only testing that one category against the baseline (Beauty). To test
whether category matters as a whole, a Type-II ANOVA F-test compares
Model 4 against the same model with `product_category` removed entirely.

```{r, echo=TRUE}
m4_no_category <- lm(log_total_amount ~ unit_price + quantity + discount_amount + age,
                      data = df)
category_anova <- anova(m4_no_category, m4_mlr_full_log)
```

```{r, echo=FALSE}
kable(category_anova, caption = "F-test: Overall Effect of Product Category on Log(Total Amount)")
```

**Findings:** Removing `product_category` significantly worsens model fit
(F = `r round(category_anova$F[2], 2)`, p < 0.001). Product category, as
a whole, is a significant predictor of spend, on top of price, quantity,
discount, and age.

## 4.3 Model Comparison

```{r, echo=FALSE}
models <- list(
  "Model 1 (SLR)"          = m1_slr,
  "Model 2 (Pricing MLR)"  = m2_mlr_pricing,
  "Model 3 (Full MLR)"     = m3_mlr_full,
  "Model 4 (Full MLR, log)"= m4_mlr_full_log
)

comparison_table <- tibble(
  Model         = names(models),
  R_Squared     = map_dbl(models, ~ round(summary(.x)$r.squared, 4)),
  Adj_R_Squared = map_dbl(models, ~ round(summary(.x)$adj.r.squared, 4)),
  RMSE          = map_dbl(models, ~ round(sqrt(mean(residuals(.x)^2)), 2))
)

kable(comparison_table, col.names = c("Model", "R2", "Adj R2", "RMSE"),
      caption = "Model fit comparison")
```

> **Note:** Models 1–3 predict `total_amount` in dollars, so their RMSE is
directly comparable. Model 4 predicts `log(total_amount)`, so its RMSE
is in log-dollar units and looks artificially small, it is not directly
comparable to Models 1–3's RMSE on the dollar scale.
Model 4 achieved the highest explanatory power (R² = 0.8551) while also satisfying regression assumptions more effectively through log transformation. Hence, Model 4 was selected as the final model.

```{r, echo=FALSE}
kable(
  tidy(m4_mlr_full_log) %>%
    select(term, estimate, std.error, statistic, p.value),
  digits = 4,
  caption = "Final Model Coefficients (Model 4)"
)
```

# 5. Diagnostic Checks

## 5.1 Multicollinearity Test (VIF)

```{r,echo=FALSE}
vif_m3 <- vif(m3_mlr_full)
vif_m4 <- vif(m4_mlr_full_log)
kable(as.data.frame(vif_m3), caption = "VIF (Model 3 and Model 4 — identical, since both use the same predictors)")
```

> All VIF values are well below the common threshold of 5, so
multicollinearity is not a concern.

## 5.2 Residual Diagnostics — Model 3 (Raw Scale)

```{r, echo=FALSE}
par(mfrow = c(2, 2))
plot(m3_mlr_full)
par(mfrow = c(1, 1))
```

> Residuals vs Fitted and Scale-Location both show a clear funnel shape,
with spread widening as fitted values increase, and the Q-Q plot bows
away from the reference line in the upper tail. These are the textbook
signs of heteroscedastic, non-normal residuals, consistent with the
right-skewed target variable identified in Section 2.4.

## 5.3 Residual Diagnostics — Model 4 (Log Scale)

```{r, echo=FALSE}
par(mfrow = c(2, 2))
plot(m4_mlr_full_log)
par(mfrow = c(1, 1))
```

> The same two plots look markedly better here: the funnel shape in
Residuals vs Fitted largely flattens out, and the Q-Q plot tracks the
reference line far more closely across its full range. This is the
direct visual confirmation that the log transformation fixed the
assumption violations seen in Model 3.

## 5.4 Assumption Summary
```{r, echo = FALSE}
assumption_table <- data.frame(
  Assumption = c(
    "Linearity",
    "Normality",
    "Homoscedasticity",
    "Multicollinearity"
  ),
  Status = c(
    "Satisfied",
    "Improved after log transformation",
    "Improved after log transformation",
    "Satisfied"
  )
)

kable(
  assumption_table,
  caption = "Regression Assumption Summary"
)
```

>Diagnostic plots indicate that the log-transformed model provides a better fit to the assumptions of linear regression. Therefore, Model 4 was selected as the final model for interpretation and hypothesis testing.

# 6. Customer Retention Analysis (H6)

>Sections 1-5 model spend as a continuous outcome. This section asks a different question, does customer identity (returning vs. new) predict anything, 

## 6.1 Two-Sample T-Test: Returning vs New Customers

> Welch two-sample t-test was used because it does not assume equal
variances between groups.

- **H6 (H0: μ_returning = μ_new):** Returning and new customers have a significant relationship with total spend.

```{r, echo=TRUE, results='hide'}
t_test_result <- t.test(total_amount ~ is_returning_customer, data = df, var.equal = FALSE)
t_test_result
```

```{r, echo=FALSE}
kable(broom::tidy(t_test_result) %>%
        select(estimate1, estimate2, statistic, p.value, conf.low, conf.high),
      digits = 3, col.names = c("Mean (Returning)", "Mean (New)", "t-stat",
                                 "p-value", "CI Low", "CI High"),
      caption = "Welch Two-Sample t-test: Total Amount by Returning-Customer Status")
```

> **Findings:** Returning customers spend $`r round(t_test_result$estimate[1], 2)`
vs. $`r round(t_test_result$estimate[2], 2)` for new customers which is not
statistically significant (t = `r round(t_test_result$statistic, 3)`,
p = `r round(t_test_result$p.value, 3)`). Return status is independent of
a single order's spend.

# 7. Hypothesis Testing Summary

```{r, echo=FALSE}
# 1. Map continuous regression predictors to their respective Hypotheses
hyp_map <- tibble(
  term = c("quantity", "unit_price", "age", "discount_amount"),
  Hypothesis = c("H1", "H2", "H3", "H4"),
  Variable = c("Quantity", "Unit Price", "Customer Age", "Discount Amount")
)

# 2. Extract results for H1 - H4 from Model 4
regression_hypothesis <- tidy(m4_mlr_full_log) %>%
  inner_join(hyp_map, by = "term") %>%
  mutate(
    Coefficient = round(estimate, 4),
    P_Value = ifelse(p.value < 0.0001, formatC(p.value, format = "e", digits = 2), as.character(round(p.value, 4))),
    Significance = ifelse(p.value < 0.05, "Significant", "Not Significant"),
    Result = ifelse(p.value < 0.05, "Supported", "Rejected")
  ) %>%
  select(Hypothesis, Variable, Coefficient, P_Value, Significance, Result)

cat_anova <- car::Anova(m4_mlr_full_log, type = "II")
cat_p_val <- cat_anova["product_category", "Pr(>F)"]

cat_hypothesis <- tibble(
  Hypothesis = "H5",
  Variable = "Product Category (Overall)",
  Coefficient = NA_real_,
  P_Value = ifelse(cat_p_val < 0.0001, formatC(cat_p_val, format = "e", digits = 2), as.character(round(cat_p_val, 4))),
  Significance = ifelse(cat_p_val < 0.05, "Significant", "Not Significant"),
  Result = ifelse(cat_p_val < 0.05, "Supported", "Rejected")
)

ttest_hypothesis <- tibble(
  Hypothesis = "H6",
  Variable = "Returning Customer Status",
  Coefficient = NA_real_,
  P_Value = ifelse(t_test_result$p.value < 0.0001, formatC(t_test_result$p.value, format = "e", digits = 2), as.character(round(t_test_result$p.value, 4))),
  Significance = ifelse(t_test_result$p.value < 0.05, "Significant", "Not Significant"),
  Result = ifelse(t_test_result$p.value < 0.05, "Supported", "Rejected")
)

hypothesis_table <- bind_rows(regression_hypothesis, ttest_hypothesis, cat_hypothesis) %>%
  arrange(Hypothesis)
kable(hypothesis_table, caption = "Summary of All Hypothesis Test Results (H1-H6)")
```


>**Interpretation (log-scale coefficients):** As the outcome for H1-H4
is `log(total_amount)`, each coefficient is approximately the
*percentage* change in spend per one-unit increase in the predictor. H5
is tested separately via the t-test in Section 6.1 and is not a
regression coefficient.

**What this means for the business:**

- **H1 (Quantity) — Supported.** Each additional unit in an order is
  associated with roughly a 55.5% increase in total spend, making order
  quantity the single strongest lever identified in this analysis.
  Bundling or minimum-order promotions could meaningfully lift revenue per
  order.
- **H2 (Unit Price) — Supported.** The per-dollar effect is small,
  meaningful mainly at large price differences which is consistent with unit
  price mattering more through category-level positioning (Section 3.4)
  than fine-grained price tweaks.
- **H3 (Customer Age) — Rejected.** Age shows no significant relationship
  with spend, so age-based targeting would not meaningfully predict order
  value.
- **H4 (Discount Amount) — Supported** Higher
  discounts are associated with slightly lower total spend on average.
  This likely reflects that heavier discounting occurs on lower-value
  orders or categories, rather than discounts actively suppressing
  spend — a correlation, not a proven causal effect.
- **H5 (Product Category) — Supported:** Expenditure differs significantly 
  across product categories. 
- **H6 (Returning vs New Customers) — Rejected.** Returning and new
  customers spend statistically indistinguishable amounts per order, so
  order value alone cannot be used to infer or predict returning-customer
  status.

# 8. Conclusion

- **H1, H2, and H5 supported:** order quantity, unit price, and product category
  are the strongest, most consistent drivers of spend.
- **H4 supported:** discount amount has a small but significant
  *negative* association with spend, contrary to the commonly assumed
  positive effect.
- **H3 and H6 rejected:** customer age has no meaningful relationship
  with spend, and returning customers do not spend significantly
  differently from new customers.
- Model 4 (log-transformed) best satisfies regression assumptions and is
  the preferred model for inference; Model 3 remains useful for
  communicating effects directly in dollars.
